### 2016-05-12 text/Chapter05.tex
Edits in implementation, consistency (Chapter 7)

Following a similar approach to Moggi, Uustalu and Vene \cite{comonads-notions} showed that
(\emph{monoidal}) \emph{comonads} uniformly capture the semantics of various kinds of context-dependent
computations~\cite{comonads-notions}. For example, data-flow computations over non-empty lists
are modelled using the non-empty list comonad $\ident{NEList}\,\alpha = \alpha + (\alpha \times \ident{NEList}\,\alpha)$.

@@

As mentioned earlier, Uustalu and Vene \cite{comonads-notions} use comonads to model data-flow
computations. They describe infinite (coinductive) streams and non-empty lists as example comonads.

\begin{example}[Non-empty list]
A non-empty list is a recursive data-type defined as $\ident{NEList}\,\alpha = \alpha + (\alpha \times \ident{NEList}\,\alpha)$.
We write \kvd{inl} and \kvd{inr} for constructors of the left and right cases, respectively. The
type \ident{NEList} forms a comonad together with the following \ident{counit} and \ident{cobind} mappings:
%
\begin{equation*}
\begin{array}{rclll}
\ident{counit}~l &\narrow{=}& h &\quad&\textnormal{when}~l=\kvd{inl}~h\\[-0.25em]
\ident{counit}~l &\narrow{=}& h &&\textnormal{when}~l=\kvd{inr}~(h, t)\\[0.5em]
\ident{cobind}\,f~l &\narrow{=}& \kvd{inl}~(f\,l) &&\textnormal{when}~l=\kvd{inl}~h\\[-0.25em]
\ident{cobind}\,f~l &\narrow{=}& \kvd{inr}~(f\,l,\;\ident{cobind}~f~t) &&\textnormal{when}~l=\kvd{inr}~(h, t)
\end{array}
\end{equation*}
\end{example}

@@

The $\ident{split}_{\cclrd{r},\cclrd{s}}$ operation splits a single comonadic value (containing a tuple)
into two separate values. Note that this does not simply duplicate the value, because the additional
context is also split. To obtain coeffects $\cclrd{r}$ and $\cclrd{s}$, the input needs to provide
\emph{at least} $\cclrd{r}$ and $\cclrd{s}$, so the tags are combined using the $\cpar$, which is often
the \emph{least upper-bound}\footnote{The $\czip$ and $\cpar$ operations are the greatest and least upper
bounds in the liveness and data-flow examples, but not for implicit parameters. However, they remain useful
as an informal analogy.}.

@@

Our last example provides the semantics of data-flow computations. The flat coeffect algebra
is formed by $(\mathbb{N}, +, \mathit{max}, \mathit{min}, 0, 0, \leq)$. In a
non-indexed version, the semantics is provided by a non-empty list. In the indexed semantics,
the index represents the number of available past values. The data type is then a pair of
the current value, followed by $n$ past values. The mappings that form the flat indexed comonad
are defined as follows:
%
\begin{equation*}
\begin{array}{l}
\ident{counit}_{0}\langle a_0 \rangle = a_0
\\[0.45em]
\ident{cobind}_{m, n}~f \langle a_0, \ldots a_{m+n} \rangle = \\[-0.25em]
\quad \langle f \langle a_0, \ldots, a_m \rangle, \ldots, f \langle a_{n}, \ldots, a_{m+n} \rangle \rangle
\\[0.45em]
\ident{merge}_{m, n} (\langle a_0, \ldots, a_m \rangle, \langle b_0, \ldots, b_n\rangle) = \\[-0.25em]
\quad \langle (a_0, b_0), \ldots, (a_{\mathit{min}(m,n)}, b_{\mathit{min}(m,n)}) \rangle
\\[0.45em]
\ident{split}_{m, n} \langle (a_0, b_0), \ldots, (a_{\mathit{max}(m,n)}, b_{\mathit{max}(m,n)}) \rangle = \\[-0.25em]
\quad (\langle a_0, \ldots, a_m \rangle, \langle b_0, \ldots, b_n\rangle)
\end{array}
\begin{array}{l}
\hspace{-2em}\ctyp{n}{\alpha} = \underbrace{\alpha \times \ldots \times \alpha}_{(n+1)-\textnormal{times}}\\[8em]
~\\
\end{array}
\end{equation*}
\end{example}

@@

The $\ident{cobind}_{m,n}$ operation requires $m + n$ elements in order to generate $n$ past results
of the $f$ function, which itself requires $m$ past values. When combining two lists,
$\ident{merge}_{m,n}$ behaves as \emph{zip} and produces a list that has the length of the shorter
argument. When splitting a list, $\ident{split}_{m, n}$ needs the maximum of the required lengths.

@@

\begin{figure*}[t]
The semantics is defined over a typing derivation:
%
\begin{equation*}
\hspace{-3em}
\begin{array}{ll}
\hspace{5.2em}\semdef
  {\coctx{\Gamma}{\cunit} \vdash x_i : \tau_i }
  {\pi_i \circ \ident{counit}_{\cunit}}
& (\emph{var})
\\[1em]
\hspace{4.9em}\semdef
  {\coctx{\Gamma}{\czero} \vdash n : \ident{num} }
  {\ident{const}~n}
& (\emph{num})
\\[1.5em]
\hspace{2.2em}\semdeff
  {\coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
  {f}
  {f \circ \ident{curry}~\ident{merge}_{\cclrd{r}, \cclrd{s}} }
& (\emph{abs})
\\[2em]
\emdefff
  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}
  {\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }
  {\begin{array}{l}\sem{\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2}\\[-0.3em]~\end{array}}
  {f}
  {g}
  {\hspace{-0.5em}\begin{array}{l}
  \ident{app}~\circ~f\hspace{-0.15em}\times\hspace{-0.15em}(\ident{cobind}_{\cclrd{s}, \cclrd{t}}~g)~\circ~\ident{split}_{\cclrd{r}, \cclrd{s} \,\cseq\, \cclrd{t}} \\[-0.3em]
  ~~\circ~\ident{map}_{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})}~\ident{dup}\\[-0.3em]
  \end{array}\hspace{-0.5em}}
& (\emph{app})
\end{array}
\end{equation*}

@@

In Section~\ref{sec:applications-flat}, we defined the semantics of concrete (flat) context-dependent
computations including implicit parameters, liveness and data-flow. Using the \emph{flat indexed
comonad} structure, we can now define a single uniform semantics that is capable of capturing all
our examples, as well as various other computations.

@@

The semantics obtained by instantiating the rules in Figure~\ref{fig:semantics-flat} with the
concrete operations defined in Example~\ref{thm:semantics-indexed-opt},
Example~\ref{thm:semantics-indexed-prod} or Example~\ref{thm:semantics-indexed-list} is the
same as the one defined in Figure~\ref{fig:applications-flat-livsem},
Figure~\ref{fig:applications-flat-implsem} and Figure~\ref{fig:applications-flat-dfsem},
respectively.
\end{theorem}
\begin{proof}
Simple expansion of the definitions for the unique typing derivation.
\end{proof}

@@

\caption{Common syntax and reduction rules of the target language}
\label{fig:semantics-target}
\end{figure}

@@

\caption{Comonadically-inspired extensions for the target language}
\label{fig:semantics-ext}
\end{figure}

@@

\caption{Additional constructs for modelling dataflow in the target language}
\label{fig:semantics-ext-df}
\end{figure}

@@

Finally, Figure~\ref{fig:semantics-ext-impl} also defines the reduction rules. The (\emph{lookup})
rule accesses an implicit parameter and (\emph{letimpl}) adds a new binding. The reduction rules
closely model the product comonad discussed in Example~\ref{thm:semantics-indexed-prod}. Reductions
for (\emph{cobind}) and (\emph{split}) restrict the set of available implicit parameters according
to the annotations and (\emph{merge}) combines them, preferring the values from the call-site.

For the semantics of implicit parameter programs that we consider, the preference of call-site
bindings over declaration-site bindings in (\emph{merge}) does not matter. The unique typing
derivations for implicit parameter coeffects obtained in Section~\ref{sec:flat-unique} always
split implicit parameters into \emph{disjoint sets}, so preferences do not come into play.

@@

\begin{figure}[h!]
\paragraph{Language syntax}

@@

\vspace{-0.5em}
\caption{Additional constructs embedding implicit parameters into the language}
\label{fig:semantics-ext-impl}
\end{figure}

@@

\begin{theorem}[Safety of context-aware langauge with implicit parameters]
  If $\Gamma \vdash e : \tau$ and $e \rightsquigarrow^* e'$ then either $e'$ is a value of type $\tau$ or
  there exists $e''$ such that $e' \rightsquigarrow e''$ and $\Gamma \vdash e'' : \tau$.
\end{theorem}
\begin{proof}
  Rule induction over $\rightsquigarrow^*$ using Theorem~\ref{thm:semantics-impl-pres} and Theorem~\ref{thm:semantics-impl-prog}.
\end{proof}

@@

In Section~\ref{sec:semantics-df} and Section~\ref{sec:semantics-proofs-impl}, we proved the
safety property of two concrete context-aware programming languages based on the coeffect
language framework. The proofs for the two systems were very similar and relied on the
same key principle.

>>>>

Following a similar approach to Moggi, Uustalu and Vene \cite{comonads-notions} showed that
(\emph{monoidal}) \emph{comonads} uniformly capture the semantics of various kinds of context-dependent
computations~\cite{comonads-notions}. For example, dataflow computations over non-empty lists
are modelled using the non-empty list comonad $\ident{NEList}\,\alpha = \alpha + (\alpha \times \ident{NEList}\,\alpha)$.

@@

As mentioned earlier, Uustalu and Vene \cite{comonads-notions} use comonads to model dataflow
computations. They describe infinite (coinductive) streams and non-empty lists as example comonads.

\begin{example}[Non-empty list]
A non-empty list is a recursive data-type defined as $\ident{NEList}\,\alpha = \alpha + (\alpha \times \ident{NEList}\,\alpha)$.
We write \kvd{inl} and \kvd{inr} for constructors of the left and right cases, respectively. The
type \ident{NEList} forms a comonad together with the following \ident{counit} and \ident{cobind} mappings:

\begin{equation*}
\begin{array}{rclll}
\ident{counit}~l &\narrow{=}& h &\quad&\textnormal{when}~l=\kvd{inl}~h\\[-0.25em]
\ident{counit}~l &\narrow{=}& h &&\textnormal{when}~l=\kvd{inr}~(h, t)\\[0.5em]
\ident{cobind}\,f~l &\narrow{=}& \kvd{inl}~(f\,l) &&\textnormal{when}~l=\kvd{inl}~h\\[-0.25em]
\ident{cobind}\,f~l &\narrow{=}& \kvd{inr}~(f\,l,\;\ident{cobind}~f~t) &&\textnormal{when}~l=\kvd{inr}~(h, t)
\end{array}
\end{equation*}
\end{example}

@@

The $\ident{split}_{\cclrd{r},\cclrd{s}}$ operation splits a single comonadic value (containing a tuple)
into two separate values. Note that this does not simply duplicate the value, because the additional
context is also split. To obtain coeffects $\cclrd{r}$ and $\cclrd{s}$, the input needs to provide
\emph{at least} $\cclrd{r}$ and $\cclrd{s}$, so the tags are combined using the $\cpar$, which is often
the \emph{least upper-bound}\footnote{The $\czip$ and $\cpar$ operations are the greatest and least upper
bounds in the liveness and dataflow examples, but not for implicit parameters. However, they remain useful
as an informal analogy.}.

@@

Our last example provides the semantics of dataflow computations. The flat coeffect algebra
is formed by $(\mathbb{N}, +, \mathit{max}, \mathit{min}, 0, 0, \leq)$. In a
non-indexed version, the semantics is provided by a non-empty list. In the indexed semantics,
the index represents the number of available past values. The data type is then a pair of
the current value, followed by $n$ past values. The mappings that form the flat indexed comonad
are defined as follows:
%
\begin{equation*}
\begin{array}{l}
\ident{counit}_{0}\langle a_0 \rangle = a_0
\\[0.45em]
\ident{cobind}_{m, n}~f \langle a_0, \ldots a_{m+n} \rangle = \\[-0.25em]
\quad \langle f \langle a_0, \ldots, a_m \rangle, \ldots, f \langle a_{n}, \ldots, a_{m+n} \rangle \rangle
\\[0.45em]
\ident{merge}_{m, n} (\langle a_0, \ldots, a_m \rangle, \langle b_0, \ldots, b_n\rangle) = \\[-0.25em]
\quad \langle (a_0, b_0), \ldots, (a_{\mathit{min}(m,n)}, b_{\mathit{min}(m,n)}) \rangle
\\[0.45em]
\ident{split}_{m, n} \langle (a_0, b_0), \ldots, (a_{\mathit{max}(m,n)}, b_{\mathit{max}(m,n)}) \rangle = \\[-0.25em]
\quad (\langle a_0, \ldots, a_m \rangle, \langle b_0, \ldots, b_n\rangle)
\end{array}
\begin{array}{l}
\hspace{-2em}\ctyp{n}{\alpha} = \underbrace{\alpha \times \ldots \times \alpha}_{(n+1)-\textnormal{times}}\\[8em]
~\\
\end{array}
\end{equation*}
\end{example}

@@

The $\ident{cobind}_{m,n}$ operation requires $m + n$ elements in order to generate $n$ past results
of $f$, which itself requires $m$ past values. When combining two lists, $\ident{merge}_{m,n}$
behaves as \emph{zip} and produces a list of length of the shorter argument. When splitting a list,
$\ident{split}_{m, n}$ needs the maximum of the lengths.

@@

\begin{figure*}[t]
The semantics is defined over a typing derivation:
%
\begin{equation*}
\hspace{-2em}
\begin{array}{ll}
\hspace{5.2em}\semdef
  {\coctx{\Gamma}{\cunit} \vdash x_i : \tau_i }
  {\pi_i \circ \ident{counit}_{\cunit}}
& (\emph{var})
\\[1em]
\hspace{4.9em}\semdef
  {\coctx{\Gamma}{\czero} \vdash n : \ident{num} }
  {\ident{const}~n}
& (\emph{num})
\\[1.5em]
\hspace{2.2em}\semdeff
  {\coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
  {f}
  {f \circ \ident{curry}~\ident{merge}_{\cclrd{r}, \cclrd{s}} }
& (\emph{abs})
\\[2em]
\emdefff
  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}
  {\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }
  {\begin{array}{l}\sem{\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2}\\[-0.3em]~\end{array}}
  {f}
  {g}
  {\hspace{-0.5em}\begin{array}{l}
  \ident{app}~\circ~f\hspace{-0.15em}\times\hspace{-0.15em}(\ident{cobind}_{\cclrd{s}, \cclrd{t}}~g)~\circ~\ident{split}_{\cclrd{r}, \cclrd{s} \,\cseq\, \cclrd{t}} \\[-0.3em]
  ~~\circ~\ident{map}_{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})}~\ident{dup}\\[-0.3em]
  \end{array}\hspace{-2em}~}
& (\emph{app})
\end{array}
\end{equation*}

@@

In Section~\ref{sec:applications-flat}, we defined the semantics of concrete (flat) context-dependent
computations including implicit parameters, liveness and dataflow. Using the \emph{flat indexed
comonad} structure, we can now define a single uniform semantics that is capable of capturing all
our examples, as well as various other computations.

@@

The semantics obtained by instantiating the rules in Figure~\ref{fig:semantics-flat} with the
concrete operations defined in Example~\ref{thm:semantics-indexed-opt},
Example~\ref{thm:semantics-indexed-prod} or Example~\ref{thm:semantics-indexed-list} is the
same as the one defined in Figure~\ref{fig:applications-flat-livsem},
Figure~\ref{fig:applications-flat-implsem} and Figure~\ref{fig:applications-flat-dfsem},
respectively.
\end{theorem}
\begin{proof}
Expansion of the definitions, using the unique typing derivation
for dataflow and liveness and any typing derivation for implicit parameters.
\end{proof}

@@

\figcaption{Common syntax and reduction rules of the target language}
\label{fig:semantics-target}
\end{figure}

@@

\figcaption{Comonadically-inspired extensions for the target language}
\label{fig:semantics-ext}
\end{figure}

@@

\figcaption{Additional constructs for modelling dataflow in the target language}
\label{fig:semantics-ext-df}
\end{figure}

@@

Finally, Figure~\ref{fig:semantics-ext-impl} also defines the reduction rules. The (\emph{lookup})
rule accesses an implicit parameter and (\emph{letimpl}) adds a new binding. The reduction rules
closely model the product comonad discussed in Example~\ref{thm:semantics-indexed-prod}. Reductions
for (\emph{cobind}) and (\emph{split}) restrict the set of available implicit parameters according
to the annotations and (\emph{merge}) combines them, preferring the values from the call site.

For the semantics of implicit parameter programs that we consider, the preference of call site
bindings over declaration-site bindings in (\emph{merge}) does not matter. The unique typing
derivations for implicit parameter coeffects obtained in Section~\ref{sec:flat-unique} always
split implicit parameters into \emph{disjoint sets}, so preferences do not come into play.

@@

\begin{figure}[h!]
\vspace{2em}
\paragraph{Language syntax}

@@

\figcaption{Additional constructs embedding implicit parameters into the language}
\label{fig:semantics-ext-impl}
\vspace{1em}
\end{figure}

@@

\begin{theorem}[Safety of context-aware language with implicit parameters]
  If $\Gamma \vdash e : \tau$ and $e \rightsquigarrow^* e'$ then either $e'$ is a value of type $\tau$ or
  there exists $e''$ such that $e' \rightsquigarrow e''$ and $\Gamma \vdash e'' : \tau$.
\end{theorem}
\begin{proof}
  Rule induction over $\rightsquigarrow^*$ using Theorem~\ref{thm:semantics-impl-pres} and Theorem~\ref{thm:semantics-impl-prog}.
\end{proof}

@@

In Section~\ref{sec:semantics-proofs-df} and Section~\ref{sec:semantics-proofs-impl}, we proved the
safety property of two concrete context-aware programming languages based on the coeffect
language framework. The proofs for the two systems were very similar and relied on the
same key principle.

### 2016-05-12 text/Chapter06.tex
Edits in implementation, consistency (Chapter 7)

\item We give a \emph{translational semantics} (Section~\ref{sec:struct-transl}) that translates
  programs from the structural coeffect calculus into a simple functional language with
  uninterpreted comonadically-inspired primitivies. We give concrete operational semantics for the
  target language for one of our sample langauges and show that well-typed programs, produced by
  translation from the coeffect calculus do not get stuck.
\end{itemize}

@@

In the structural coeffect calculus, the lambda abstraction rule does not introduce ambiguity
in the typing. This is in contrast with flat coeffect systems (most importantly, the one for
implicit parameters), where lambda abstraction allowed arbitrary splitting of context requirements.
In structural coeffect systems, the context requirements placed on the call-site (attached to
the function type) are those of the bound variable.

@@

\paragraph{Expressions.}
A semantics of structural coeffect calculi is shown in Figure~\ref{fig:struct-semantics}. The
semantics is written as composition of morphisms using a number of auxiliary definitions. Due to
the eqivalence between Cartesian Closed Categories and the $\lambda$-calculus, we will treat it
as specifying translation to a target functional langauge in Section~\ref{sec:struct-transl}.

@@

The semantics obtained by instantiating the rules in Figure~\ref{fig:struct-semantics} with the
concrete operations defined in Example~\ref{thm:struct-live},
Example~\ref{thm:struct-dfl} or Example~\ref{thm:struct-bll} is the
same as the one defined in Figure~\ref{fig:applications-struct-livesem},
Section~\ref{sec:applications-structural-dataflow} and Section~\ref{sec:applications-struct-bll},
respectively.
\end{theorem}
\begin{proof}
Simple expansion of the definitions for the unique typing derivation.
\end{proof}

@@

This section mirrors the development presented in Chapter~\ref{ch:semantics} for flat coeffect
calculus. We extend the simple target language with additional constructs inspired by structural
indexed comonads (Section~\ref{sec:struct-transl-ext}), define the translation to the target
langauge (Section~\ref{sec:struct-transl-transl}) and prove safety of one sample language
(Section~\ref{sec:struct-transl-safety}) -- we choose structural dataflow to allow easy
comparison with the flat system.

@@

\caption{Comonadically-inspired extensions for structural coeffects}
\label{fig:struct-semantics-ext}
\end{figure}

@@

The target language with comonadically-inspired primitives provides a framework that can be used
to model a variety of structural context-aware langauges and prove their safety. As outlined in
Section~\ref{sec:semantics-generalising}, the key principle that guarantees the safety of the
target language is a correspondence between coeffect annotations and the values they represent.
In a more expressive target language (such as Haskell or Agda), it would be sufficient to
provide an implementation of the comonadically-inspired primitives for the concrete domain-specific
language.

@@

\caption{Additional constructs for modelling structural dataflow}
\label{fig:structural-ext-df}
\end{figure}

@@

The Figure~\ref{fig:structural-ext-df} extends the target functional langauge with constructs,
typing rules, translation rules and reduction rules needed for modelling structural dataflow.
In the structural model, the type $\ctyp{\alift{\cclrd{r_0}, \ldots, \cclrd{r_n}}}{(\tau_0\xtimes\ldots\xtimes\tau_n)}$
represents a structure that provides values and additional contexts for variables of types
$\tau_0,\ldots,\tau_n$ with contextual capabilities as specified by corresponding coeffect annotations
$\cclrd{r_0}, \ldots, \cclrd{r_n}$.

>>>>

\item We give a \emph{translational semantics} (Section~\ref{sec:struct-transl}) that translates
  programs from the structural coeffect calculus into a simple functional language with
  uninterpreted comonadically-inspired primitivies. We give concrete operational semantics for the
  target language for one of our sample languages and show that well-typed programs, produced by
  translation from the coeffect calculus do not get stuck.
\end{itemize}

@@

In the structural coeffect calculus, the lambda abstraction rule does not introduce ambiguity
in the typing. This is in contrast with flat coeffect systems (most importantly, the one for
implicit parameters), where lambda abstraction allowed arbitrary splitting of context requirements.
In structural coeffect systems, the context requirements placed on the call site (attached to
the function type) are those of the bound variable.

@@

\paragraph{Expressions.}
A semantics of structural coeffect calculi is shown in Figure~\ref{fig:struct-semantics}. The
semantics is written as composition of morphisms using a number of auxiliary definitions. Due to
the eqivalence between Cartesian Closed Categories and the $\lambda$-calculus, we will treat it
as specifying translation to a target functional language in Section~\ref{sec:struct-transl}.

@@

The semantics obtained by instantiating the rules in Figure~\ref{fig:struct-semantics} with the
concrete operations defined in Example~\ref{thm:struct-live},
Example~\ref{thm:struct-dfl} or Example~\ref{thm:struct-bll} is the
same as the one defined in Figure~\ref{fig:applications-struct-livesem},
Section~\ref{sec:applications-structural-dataflow} and Section~\ref{sec:applications-struct-bll},
respectively.
\end{theorem}
\begin{proof}
Expansion of the definitions for the unique typing derivation.
\end{proof}

@@

This section mirrors the development presented in Chapter~\ref{ch:semantics} for flat coeffect
calculus. We extend the simple target language with additional constructs inspired by structural
indexed comonads (Section~\ref{sec:struct-transl-ext}), define the translation to the target
language (Section~\ref{sec:struct-transl-transl}) and prove safety of one sample language
(Section~\ref{sec:struct-transl-safety}) -- we choose structural dataflow to allow easy
comparison with the flat system.

@@

\figcaption{Comonadically-inspired extensions for structural coeffects}
\label{fig:struct-semantics-ext}
\end{figure}

@@

The target language with comonadically-inspired primitives provides a framework that can be used
to model a variety of structural context-aware languages and prove their safety. As outlined in
Section~\ref{sec:semantics-generalising}, the key principle that guarantees the safety of the
target language is a correspondence between coeffect annotations and the values they represent.
In a more expressive target language (such as Haskell or Agda), it would be sufficient to
provide an implementation of the comonadically-inspired primitives for the concrete domain-specific
language.

@@

\figcaption{Additional constructs for modelling structural dataflow}
\label{fig:structural-ext-df}
\end{figure}

@@

The Figure~\ref{fig:structural-ext-df} extends the target functional language with constructs,
typing rules, translation rules and reduction rules needed for modelling structural dataflow.
In the structural model, the type $\ctyp{\alift{\cclrd{r_0}, \ldots, \cclrd{r_n}}}{(\tau_0\xtimes\ldots\xtimes\tau_n)}$
represents a structure that provides values and additional contexts for variables of types
$\tau_0,\ldots,\tau_n$ with contextual capabilities as specified by corresponding coeffect annotations
$\cclrd{r_0}, \ldots, \cclrd{r_n}$.
